### 2022-11-28 response.tex

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript. 
We now discuss potential mechanisms driving shorter generation intervals of the Omicron variant in detail.
In doing so, we distinguish intrinsic generation intervals, which represent changes in infectiousness over the course of infection, from realized generation intervals, which represent time between actual transmission events---our estimates correspond to the latter quantity.
We also discuss the uncertainties associated with possible effects of immunity on the generation-interval distributions of the Omicron and Delta variants.
As the reviewer pointed out, it is true that immune evasiveness of the Omicron variant is the main explanation for the population-level success of Omicron---however, this mechanism does not require or depend on changes in generation-interval distributions as we explain.

Below please find our detailed responses to reviewers.

@@

\revtext{Less attention is paid to the more important question of WHY the omicron variant had such a short generation time in this setting.}

We have now added text to the discussion explaining different types of generation intervals (intrinsic vs realized) and why the realized generation interval of the Omicron variant might be shorter:

``Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
\cite{ito2022estimating,selby2022generation} estimated the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
The realized generation intervals represent time between actual infection events, and are different from the intrinsic generation intervals, which reflect the average profile of infectiousness of infected individuals over the course of their infections \citep{champredon2015intrinsic}.

Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals.
For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} may lead to faster recovery, which could in turn shorten the intrinsic generation interval \citep{roberts2007model}.
Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals).
In our analysis, a greater proportion of Omicron than Delta infections would have been in people with pre-existing immunity (because these people are less likely to be infected by Delta);
this allowed the Omicron variant to rapidly replace the previously dominant Delta variant at the population level.
However, the individual-level relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is difficult to predict given its high immune evasiveness:
some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, do not depend on these possible changes in the generation-interval distribution:
immune evasion could have allowed Omicron to replace Delta even without differences in intrinsic generation-interval distributions.

Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions (e.g., a lack of changes in viral kinetics or the effect of pre-existing immunity), their realized generation-interval distributions could still differ.
For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, could also lead to shorter realized generation intervals.
However, this explanation is also less likely given similarities in the inferred incubation-period distributions.

Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized. 
To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simple example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, leading to shorter realized generation intervals.
Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).

Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but the difference between the intrinsic infectiousness profiles remains unclear.
In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their infectiousness profiles may be smaller than the estimated differences in their realized generation-interval distributions.
In addition, the counter-factual initial intrinsic generation intervals (for an immunologically naive population) of both variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating initial intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
and can therefore be expected to improve estimates of effective reproduction numbers.''

@@

\revtext{The authors discuss the possibility of a "network effect" which is predicated on omicron having a higher Ro and therefore somewhat circular.}

We don't believe that this explanation is circular, and have tried to clarify our explication further. As we explain in the Discussion, the success of the Omicron variant at the population level is due to immune evasion (also pointed out by reviwer 1), and this doesn't require shorter generation intervals. We now provide a concrete example for the network effect, explaining how higher reproduction number can lead to shorter \emph{realized} generation intervals even when two variants have identical \emph{intrinsic} generation intervals:

``Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized. 
To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simple example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, leading to shorter realized generation intervals.
Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).''

@@

We now discuss the possible effects of immune evasion more thorougly.

\revtext{To address this properly, the authors will have to account for the different proportions of the population currently immune to infection by either variant and how the level of cross-immunity between them affects our observations of transmission advantage. The authors should at least justify why they have not taken such an obvious feature of this multi-strain system into account.}

While we agree that immune evasion is the most likely explanation for the success of omicron in almost every global setting, its success or immune evasion does not require or depend on changes in generation-interval distributions.
We now provide a more thorough explanation for how pre-existing immunity and immune evasiveness of the Omicron variant might shape the generation-interval distribution:

``Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals).
In our analysis, a greater proportion of Omicron than Delta infections would have been in people with pre-existing immunity (because these people are less likely to be infected by Delta);
this allowed the Omicron variant to rapidly replace the previously dominant Delta variant at the population level.
However, the individual-level relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is difficult to predict given its high immune evasiveness:
some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, do not depend on these possible changes in the generation-interval distribution:
immune evasion could have allowed Omicron to replace Delta even without differences in intrinsic generation-interval distributions.''

### 2022-11-28 omicron-generation.tex
done

Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals.
For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} may lead to faster recovery, which could in turn shorten the intrinsic generation interval \citep{roberts2007model}.
Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals).
In our analysis, a greater proportion of Omicron than Delta infections would have been in people with pre-existing immunity (because these people are less likely to be infected by Delta);
this allowed the Omicron variant to rapidly replace the previously dominant Delta variant at the population level.
However, the individual-level relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is difficult to predict given its high immune evasiveness:
some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasiveness may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, are separate from possible changes in the generation-interval distribution that we have outlined above.
Even if Delta and Omicron variants had identical intrinsic generation-interval distributions, immune evasiveness alone would allow the Omicron variant to become dominant.
Therefore, we might expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections to be more strongly influenced by the intrinsic differences in their viral kinetics.

Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions, their realized generation-interval distributions could still differ.
For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, could also lead to shorter realized generation intervals.
However, this explanation is also less likely given similarities in the inferred incubation-period distributions.

Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized. 
To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simplest example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).

Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but the difference between the intrinsic infectiousness profiles remains unclear.
In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
and can therefore be expected to improve estimates of effective reproduction numbers.

>>>>

Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals.
For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} may lead to faster recovery, which could in turn shorten the intrinsic generation interval \citep{roberts2007model}.
Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals).
In our analysis, a greater proportion of Omicron than Delta infections would have been in people with pre-existing immunity (because these people are less likely to be infected by Delta);
this allowed the Omicron variant to rapidly replace the previously dominant Delta variant at the population level.
However, the individual-level relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is difficult to predict given its high immune evasiveness:
some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, do not depend on these possible changes in the generation-interval distribution:
immune evasion could have allowed Omicron to replace Delta even without differences in intrinsic generation-interval distributions.

Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions (e.g., a lack of changes in viral kinetics or the effect of pre-existing immunity), their realized generation-interval distributions could still differ.
For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, could also lead to shorter realized generation intervals.
However, this explanation is also less likely given similarities in the inferred incubation-period distributions.

Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized. 
To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simple example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, leading to shorter realized generation intervals.
Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).

Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but the difference between the intrinsic infectiousness profiles remains unclear.
In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their infectiousness profiles may be smaller than the estimated differences in their realized generation-interval distributions.
In addition, the counter-factual initial intrinsic generation intervals (for an immunologically naive population) of both variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating initial intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
and can therefore be expected to improve estimates of effective reproduction numbers.
